---
title: "Spatial Modeling of Cardiovascular Disease Incidence Positively Associated with PM2.5"
format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: true
    doublespacing: true
    runninghead: "CVD Related to PM2.5"
  arxiv-html: default
author:
    
  - name: Johan Booc
    affiliations:
      - name: Texas A\&M University
        department: Department of Statistics
    email: jbooc24@tamu.edu
  - name: Christina Kim
    affiliations:
      - name: Texas A\&M University
        department: Department of Statistics
    email: christinaykim3@tamu.edu 
  - name: Shombit Roy
    affiliations:
      - name: Texas A\&M University
        department: Department of Statistics
    email: shombit123@tamu.edu
abstract: |
  Cardiovascular disease (CVD) is the leading cause of death in the United States. By using a spatial modeling technique (geographically weighted regression), we found the concentration of PM 2.5 is centered in the Stroke Belt region. Policymakers and health practitioners can use these results to identify targeted interventions to curb the increasing rates of CVD, aiming to halt one of the world’s deadliest diseases.  
keywords: 
  - Fine Particulate Matter (PM2_5)
  - Cardiovascular Disease (CVD)
  - Cardiovascular Mortality (CVM)
bibliography: bibliography.bib  
editor: 
  markdown: 
    wrap: 72
---

# Introduction {#sec-intro}

Broad strokes: Explain how cardiovascular disease is the leading cause
of death in the US, our paper filing the gap by putting the focus on the
18-44-year-old age group —\> As cardiovascular disease is the leading
cause of death in the US, numerous past studies have been done on this
disease, specifically on the older population, the 65+ year ag-group.
Our study aims to further investigate the impact, specifically on the
18-44-year-old age group, as there has been less interest in the effects
of CVD for this age population.

Specifics: Explicitly state how the CVD incidence rate varies spatially
in the US, for our covariates. Our approach involves analyzing this
relationship to produce risk prediction estimates for our specified age
group. —\> There are many factors involved in influencing the incidence
rate of CVD, ranging from genetics, lifestyle, diet, and smoking. We
look into the covariates - PM 2.5 concentration, median household
income, and unemployment rates - and analyze the spatial variation of
them. By doing so, we can construct a risk prediction model for our
specified age population, to decrease future CVD incidence rates.

Central thesis: Use a geographically weighted model approach to produce
visualizations and analyze the relationship between our response
variable (CVD deaths) and its covariates (air pollutant concentration
and socioeconomic factors), specifically focusing on spatial variation.
—\> The geographically weighted regression model uses local variables
and weights, allowing us to produce spatial visualizations of the
regional variations in the US between our response variable (CVD deaths)
and its covariates. Other methods such as the traditional linear
regression model and clustering have been used in past CVD studies,
however, we found the GWR model to be more efficient in analyzing the
dynamic relationship of CVD and the factors that contribute to it.

Step back: The overall outcome of our study involves aiding policymakers
and health practitioners develop the necessary interventions in the
targeted regions that are most affected by CVD. —\> By using a
geographically weighted regression approach and constructing a risk
prediction model, the outcome of our study aids policymakers and health
practitioners in implementing the necessary interventions for each
targeted region most affected by CVD. Tailoring to each specific region
of the US first makes it easier to seemingly decrease CVD incidence
rates on a national scale.

# Related Works

Paragraph 1. (Review of regression approaches in cardiovascular disease
mortality) \>\>\>\> There's been a growing popularity for the use of
spatial models in the epidemiological domain to analyze the factors and
spatial distribution of a certain disease. The spatial models and
techniques used differ between studies however the end goal remains the
same, reduce the disease mortality rates for the population. Statistical
regression models have been a popular choice, specifically for
CVD-related studies. The factors that compromise the underlying causes
for CVD can be seen as a dynamic and interconnected web.
@zelkoGeographicallyWeightedModeling2023 examines the relationship
between CVD and covariates similar to our study, such as air pollution,
social determinants, and county-level data. By using a GWR model, they
found a correlation between household income, race, and healthcare
access. Their results also showed that counties in the South had the
highest PM 2.5 concentrations.

Paragraph 2. (Review of geographically weighted regression models)
\>\>\>\> The geographically weighted regression model (GWR) stands out
from the traditional regression models to explore spatial data by doing
so on the local scale as well as taking into account varied coefficients
for a certain spatial unit (Gebread et. al). The GWR model uses the
weighted least square method to estimate the regression coefficients. In
general, we are interested in seeing values closer to the point of
interest than values farther from it, as they carry more weight and have
a greater influence. The GWR model includes the parameter, neighborhood
(also known as the bandwidth).

Paragraph 3. (Direct comparison with OLS approaches) \>\>\>\> As seen in
other studies, the Ordinary Least Squares model is a popular method for
public health studies. Compared to the GWR model, the OLS model
generates the regression coefficients on a global scale. Because of
this, the OLS method may not be the ideal choice. as it is prone to
hidden spatial variability. However, our paper takes into account the
fact that socioeconomic factors are not constant on a time and space
basis with cardiovascular disease by using a GWR model, as it's a much
more complex relationship. That's not to say the OLS model has no use in
epidemiological research studies. One of the main goals of public health
is to better the population as a whole. The OLS model is efficient in
capturing the average differences between covariates. To get to the root
cause of CVD mortality rates, we need to start with the GRW model, to
see overall improvement in the OLS model later on.

Paragraph 4. (GWR with CVD) \>\>\>\> The GWR model shows whether or not
a linear relationship exists between cardiovascular disease deaths and
the covariates, making it suitable to see where the areas of focus
should be placed in terms of improvement as well as an increase presence
of healthcare practitioners. The GWR model is better for analyzing
disparities at the local-scale, see which socioeconomic factor affects
different regions, and then implement a policy for that specific region.
Past studies tend to soley focus on the trends of socioeconomic
covariates and its spatial patterning. Our paper expands on past
studies, by including the concentration of air pollutants as one of our
covariates. By doing so, we are able to fully explore the dynamic
relationship between geographical units and CVD incidence/mortality
rates.

Paragraph 5. (Preventative risk measures + impact) \>\>\>\> Risk
assessment and risk estimates helps to understand the key factors
associated with CVD. The concentration of specific races vary by
counties so by studying the underlying socioeconomic covariates of CVD
mortality on the county-level, researchers and health practitioners are
thus able to curate the necessary risk preventative measures and
allocate resources to the areas that need it the most. Furthermore, the
dose-response relationship of air pollutant concentrations in a region
and its effects can be analyzed. By putting the focus on smaller units,
CVD-mortality rates will start to see a decline, nationally and
globally. \>\>\>\>\>\>\> a7efe0bd48d8a3b96a6a1f7afe611b016c8972ad

# Methods

## Data Collection

The
"Center_for_Medicare\_**Medicaid_Services\_*CMS***\_Medicare_Claims_data_20240208.csv"
file from (CDC) website was loaded to analyze Medicare claims data,
specifically Cardiovascular death rates across different counties and
States in the US. Racial and geographic data were retrieved from the
Vacdata in R. The script "Code to get ACS Socioeconomic file" from the
'functions' folder in the linked GitHub repository was used to extract
median income data. Air quality data was sourced from the
"pm25-2015.shp" file, and unemployment rates for the year 2015 were
gathered using the "unemploymentRate2015.csv" file.

## Data Preprocessing

Several steps were taken to ensure the reliability and accuracy of the
results when preparing the data for statistical analysis. The data from
various sources was integrated into a final shapefile named
"finalCVDshapes.shp." This integration was achieved through coding in R,
specifically focusing on data from 2015, which allowed for a consistent
time frame across all data sets. During the cleaning and processing
phase, features with empty geometries were removed from the shapefile to
ensure the removal of NA values in the dataset. Missing data entries
were also omitted, which led to one county’s data being excluded.
Centroids of the multipolygon geometries were calculated to have spatial
analysis by providing a single reference point for each region. This
step was essential for conducting precise location-based assessments.
Lastly, the data frame was transformed into a SpatialPolygonsDataFrame,
making it suitable for use in the Geographic Weighted Regression (GWR)
model. This transformation was crucial as it enabled the analysis to
incorporate spatial relationships within the data.

## Statistical Analysis

In the statistical analysis, a Geographically Weighted Regression (GWR)
model was used and ran using several R packages, including sf, GWmodel,
ggplot2, and tidyverse. These packages were integral in setting up the
model framework and for executing the analysis, providing tools for
spatial data manipulation, regression modeling, and data visualization.
The optimal bandwidth for the GWR model was determined using the bw.gwr
function, which employed a Gaussian kernel. The reason Gaussian kernel
was used was because it transforms the input data points into a
higher-dimensional space, where the similarity between two data points
is measured as the Euclidean distance between them. This function
modeled the response variable, Cardiovascular death rate per
100000(Data_Vl), as a function of racial percentages, air pollutant
concentrations, estimated median income, and unemployment rates for the
year 2015. Optimizing the bandwidth is crucial as it affects the model's
sensitivity to local variations, enhancing its accuracy in reflecting
spatial heterogeneity. Parallel processing was employed (using the
parallel.method = "omp" setting at default). The GWR itself was
performed using the identified optimal bandwidth, incorporating the same
predictors as those used for bandwidth selection. This ensures that the
model's findings are reliable and applicable to the predictors. Finally,
the GWR results were converted back into an sf object for visualization
purposes. This conversion is critical for effectively allowing the
covariates of the regression outcomes to be plotted and interpreted
visually.

## Mapping

Generated maps using ggplot to visualize the spatial distribution of the
dependent variable (deaths from CVD per 100,000 people) across different
counties in the US, highlighting the significance of the variables, and
plotted the significance of each variable within different regions to
assess the overall effect on the CVD death rate.

In conclusion the approach allows for examining how various
socioeconomic, environmental, and demographic factors influence health
outcomes across different regions in the States. The use of GWR helps
understand local variations in these factors' impacts.

This methodological outline ensures that each step of the data handling,
analysis, and visualization process is documented, providing
transparency and reproducibility of the research findings.

```{r include=FALSE, cache=FALSE}
source("./Functions/gwmodelJohan.R")
source("./Functions/GWmodelRscript.R")
```

```{r}
gwr_results_sf%>%ggplot() +
  geom_sf(aes(fill = gwr_results_sf$y),
          color = scales::alpha("black",
                                alpha = 0.1
                                )) +
  scale_fill_gradientn(colours = terrain.colors(8),
                       #limits = c(19,530),
                       breaks = c(50,150,300)) +
  theme(text = element_text(size = 10), 
        legend.position = "bottom") + 
  labs(title = "CVD Deaths, 2015",
       caption = "Fig 1",
       fill = "Deaths from CVD per 100,000 people")
```

# Results

```{r}
gwr_results
```

## Table Summary

The analysis compares two statistical models to explore the
relationships between socio-economic, demographic, and environmental
variables and death rates. The first is a global regression model that
does not consider spatial correlation despite revealing that all
predictors are highly statistically significant. Hence, while the model
does suggest that our variables are indeed important, the global model
may overlook local variations that are crucial in understanding the true
nature of the data.

On the other hand, the geographically weighted regression (GWR) model
incorporates spatial variation, which is a critical factor given the
context of the data. We can see how well the GWR model worked by looking
at the R-squared value, which is .870, a significant improvement over
the global model. This R-squared value, along with the use of spatial
statistics, shows the non-uniform relationship between the predictors
and the response variable across different geographical areas. In
summary, by accommodating the spatial component present in the data, the
GWR model provides a more realistic interpretation of how various
factors influence death rates across different regions.

## Plots

In Figure #1 the plot represents the results of Geographically Weighted
Regression (GWR) analysis of particulate concentrations and their
correlation with cardiovascular disease (CVD) death rates in 2015. The
particulate concentrations include all the covariates mentioned in the
methods section. The regions along the higher latitudes (nearing 45°N)
and towards the eastern section (approaching 80°W) display darker
shades, suggesting higher CVD death rates in these areas, specifically
the midwest and southwest regions.

Figure #2 represents the local significance and magnitude of the '%
White' demographic parameter on cardiovascular disease outcomes. From
the plot, the prominent dark purple areas in the central part of the
country, extending towards the southeastern regions, indicate a
significant negative correlation between the percentage of the white
population and CVD outcomes in these areas. Same with the Northwest
region, however, there are few areas, particularly in the southwest near
New Mexico and Arizona, where there are positively correlated to CVD
death rate, with the white regions of counties suggesting no
significance.

Figure #3 represents the local significance and magnitude of the '%
African' demographic parameter on cardiovascular disease outcomes. From
the plot, areas located approximately in the northern central region
indicate a significant positive correlation to CVD rate, and areas,
notably in the central to southeastern areas of the map, suggest a
significant negative correlation, with the white regions of counties
suggesting no significance.

Figure #4 represents the local significance and magnitude of the '%
Hispanic' demographic parameter on cardiovascular disease outcomes. From
the plot, areas across the central and southeastern regions are shaded
in purple, suggesting a significant negative correlation between the
percentage of the Hispanic population and CVD outcomes. There are
isolated red patches in the north-central region, indicating areas where
an increased percentage of the Hispanic population correlates with
higher CVD outcomes. Lastly, as per the last figure, the white regions
are specified as having no significance.

Figure #5 represents the local significance and magnitude of the '%
Asian' demographic parameter on cardiovascular disease outcomes. From
the plot, a substantial portion of the map, particularly across the
central to eastern regions, is colored in various shades of purple. This
suggests that in these areas, an increased percentage of the Asian
population correlates with lower CVD outcomes. However, there are some
parts of the West where higher percentages of the Asian population are
associated with an increase in CVD outcomes.

Figure #6 represents the local significance and magnitude of the “%p2.5
air quality” parameter on cardiovascular disease outcomes. From the
plot, in the southern and northeastern regions, there increased levels
of PM2.5 colored as red in the plot, meaning they are associated with
higher rates of CVD. There are some regions in the West where it was
negatively correlated with CVD. As usual, the white areas signify no
significance.

Figure #7 represents the local significance and magnitude of the “median
income” parameter on cardiovascular disease outcomes. From the plot,
most of the region suggests that there is significant negative
correlation between the median income parameter and CVD outcomes.
However, some parts of Texas indicate a positive correlation to CVD
outcomes. As usual, the white areas signify no significance.

Figure #8 represents the local significance and magnitude of the
“Unemployment” parameter on cardiovascular disease outcomes. From the
plot, in most of the central region, there is a positive correlation to
CVD rates, and in the southwest region, you can see a negative
correlation to CVD rates. As usual, the white areas signify no
significance.

```{r}
#| layout-ncol: 2
#| layout-nrow: 2
#significance
local_t_values <- gwr_results_sf$prc_wht_TV
local_p_values <- 2 * pt(-abs(local_t_values), df = 2480.149) #df number cam from the Gw diagnostic


gwr_results_sf$prc_wht_p_values <- local_p_values


gwr_results_sf$prc_wht_signif_coef <- ifelse(gwr_results_sf$prc_wht_p_values < 0.05, 
                                             gwr_results_sf$prc_wht, 
                                             NA)  # Set to NA if not significant

library(ggplot2)
ggplot(gwr_results_sf) +
  geom_sf(aes(fill = prc_wht_signif_coef * gwr_results_sf$prc_wht/10000), color = alpha("black",0.35), size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                      name = "Magnitude", limits = NULL) +  
  labs(title = "Local Significance and Magnitude of '% White' Parameter",
       subtitle = "Geographically Weighted Regression (GWR)",
       caption = "Fig 2: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "right")

#significance
local_t_values_2 <- gwr_results_sf$prc_frc_TV
local_p_values_2 <- 2 * pt(-abs(local_t_values_2), df = 2480.149) #df number cam from the Gw diagnostic


gwr_results_sf$prc_af_p_values <- local_p_values_2


gwr_results_sf$prc_af_signif_coef <- ifelse(gwr_results_sf$prc_af_p_values < 0.05, 
                                             gwr_results_sf$prc_frc, 
                                             NA)  # Set to NA if not significant

library(ggplot2)
ggplot(gwr_results_sf) +
  geom_sf(aes(fill = prc_af_signif_coef * gwr_results_sf$prc_frc/10000), color = alpha("black",0.35), size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                      name = "Magnitude", limits = NULL) +  
  labs(title = "Local Significance and Magnitude of '% African American' Parameter",
       subtitle = "Geographically Weighted Regression (GWR)",
       caption = "Fig 3: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "right")

#significance
local_t_values_3 <- gwr_results_sf$prc_hsp_TV
local_p_values_3 <- 2 * pt(-abs(local_t_values_3), df = 2480.149) #df number cam from the Gw diagnostic



gwr_results_sf$prc_hsp_p_values <- local_p_values_3


gwr_results_sf$prc_hsp_signif_coef <- ifelse(gwr_results_sf$prc_hsp_p_values < 0.05, 
                                             gwr_results_sf$prc_hsp, 
                                             NA)  # Set to NA if not significant

library(ggplot2)
ggplot(gwr_results_sf) +
  geom_sf(aes(fill = prc_hsp_signif_coef * gwr_results_sf$prc_hsp/10000), color = alpha("black",0.35), size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                      name = "Magnitude", limits = NULL) +  
  labs(title = "Local Significance and Magnitude of '% Hispanic' Parameter",
       subtitle = "Geographically Weighted Regression (GWR)",
       caption = "Fig 4: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "right")

#significance
local_t_values_4 <- gwr_results_sf$perc_sn_TV
local_p_values_4 <- 2 * pt(-abs(local_t_values_4), df = 2480.149) #df number cam from the Gw diagnostic


gwr_results_sf$prc_sn_p_values <- local_p_values_4


gwr_results_sf$prc_sn_signif_coef <- ifelse(gwr_results_sf$prc_sn_p_values < 0.05, 
                                             gwr_results_sf$perc_sn, 
                                             NA)  # Set to NA if not significant

library(ggplot2)
ggplot(gwr_results_sf) +
  geom_sf(aes(fill = prc_sn_signif_coef * gwr_results_sf$perc_sn/10000), color = alpha("black",0.35), size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                      name = "Magnitude", limits = NULL) +  
  labs(title = "Local Significance and Magnitude of '% Asian' Parameter",
       subtitle = "Geographically Weighted Regression (GWR)",
       caption = "Fig 5: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "right")
```

-   Demographic Impact: The percentage of white and Hispanic populations
    shows significant regional variations in association with death
    rates. Higher proportions of white populations correlate with lower
    death rates in central areas, whereas higher percentages of Hispanic
    populations are linked to lower death rates in the West and
    Southwest. Conversely, higher percentages of African American
    populations are associated with higher death rates in certain
    Midwestern and Southeastern regions.

```{r}
#significance
local_t_values_5 <- gwr_results_sf$p2_5_20_TV
local_p_values_5 <- 2 * pt(-abs(local_t_values_5), df = 2480.149) #df number came from the Gw diagnostic



gwr_results_sf$prc_p2_p_values <- local_p_values_5


gwr_results_sf$prc_p2_signif_coef <- ifelse(gwr_results_sf$prc_p2_p_values < 0.05, 
                                             gwr_results_sf$p2_5_20, 
                                             NA)  # Set to NA if not significant

library(ggplot2)
ggplot(gwr_results_sf) +
  geom_sf(aes(fill = prc_p2_signif_coef), color = "black", size = 0.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                      name = "Coefficient Magnitude", limits = NULL) +  
  labs(title = "Local Significance and Magnitude of PM2.5 Parameter",
       subtitle = "Geographically Weighted Regression (GWR)",
       caption = "Fig 6: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "right")
```

-   Environmental Influence: Air quality, indicated by PM2.5 levels,
    demonstrates a significant positive relationship with death rates,
    particularly east of the Rockies, highlighting environmental health
    as a major concern.

```{r}
#significance
    local_t_values_6 <- gwr_results_sf$estimat_TV
    local_p_values_6 <- 2 * pt(-abs(local_t_values_6), df = 2480.149) #df number cam from the Gw diagnostic


    gwr_results_sf$prc_est_p_values <- local_p_values_6


    gwr_results_sf$prc_est_signif_coef <- ifelse(gwr_results_sf$prc_est_p_values < 0.05, 
                                                 gwr_results_sf$estimat, 
                                                 NA)  # Set to NA if not significant

    library(ggplot2)
    ggplot(gwr_results_sf) +
      geom_sf(aes(fill = prc_est_signif_coef * log(merge2015GW$estimat)), color = "black", size = 0.1) +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                          name = "Coefficient Magnitude", limits = NULL) +  
      labs(title = "Local Significance of Median Income Parameter",
           subtitle = "Geographically Weighted Regression (GWR)",
           caption = "Fig 7: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
      theme_minimal() +
      theme(text = element_text(size = 12), legend.position = "right")

    #significance
    local_t_values_7 <- gwr_results_sf$U__2015
    local_p_values_7 <- 2 * pt(-abs(local_t_values_7), df = 2480.149) #df number cam from the Gw diagnostic


    gwr_results_sf$prc_U_2015_p_values <- local_p_values_7


    gwr_results_sf$prc_U_2015_signif_coef <- ifelse(gwr_results_sf$prc_U_2015_p_values < 0.05, 
                                                 gwr_results_sf$U__2015, 
                                                 NA)  # Set to NA if not significant

    library(ggplot2)
    ggplot(gwr_results_sf) +
      geom_sf(aes(fill = prc_U_2015_signif_coef), color = "black", size = 0.1) +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", na.value = "white", 
                          name = "Coefficient Magnitude", limits = NULL) +  
      labs(title = "Local Significance of Unemployment Parameter",
           subtitle = "Geographically Weighted Regression (GWR)",
           caption = "Fig 8: Blue indicates lower magnitude of effect;\n Red indicates larger magnitude of effect;\nWhite indicates no significant effect") +
      theme_minimal() +
      theme(text = element_text(size = 12), legend.position = "right")
```

<!-- -->

-   Socio-economic Correlation: Median income levels across many regions
    show a consistent negative association with death rates, suggesting
    that higher income areas generally experience fewer deaths.

# Discussion

The purpose of this study was to fill a gap in prior studies where the
impacts of CVD were primarily studied within the Stroke Belt, rather
than the country at large.

Our goal was to put our findings towards answering the following
question: what are the socioeconomic and environmental factors affecting
CVD rates in the United States?

These maps reveal that the relationships between race, socio-economic
factors, environmental quality, and death rates are complex and highly
localized. The significance and strength of these relationships vary
considerably across different parts of the United States. In contrast,
some of the socio-economic factors such as median income show
widespread, consistent significance, implying that the significance of
the relationship with CVD outcomes is nearly constant across various
locations. This differs from the heterogeneous local significance that
we observed in the majority of our other variables.

Another factor to consider from the plots is the percentage of each race
that inhabits each county. We can observe that for African Americans,
Hispanics, and Asians that the impacts are significant in localized
areas of the country, which may reflect underlying health disparities in
access to medical care.

-   One limitation of race percentages is the misreporting of medical
    records affecting minorities (Tabb et al.) However, this oversight
    lends further credence to the fact that intervention is needed in
    order to combat racial health disparities. We can look at the
    Variance Inflation Factor (VIF) to show that the multicollinearity
    does not impact our results:

```{r eval = FALSE}
dp = cbind(finalCVD$lon, finalCVD$lat)
DM = gw.dist(dp.locat = dp)
#collin = gwr.collin.diagno(Data_Vl ~ 
                           # prc_wht + prc_frc + prc_hsp + perc_sn + 
                          #  p2_5_20 + estimat + U__2015, 
                          #data = mergedf_spatial, 
                          #bw = opt_bandwidth, 
                          #kernel = "gaussian", 
                          #adaptive = FALSE, 
                          #dMat = DM)
```

P3: Air quality has a broad impact, suggesting environmental health
concerns that might require region-specific policies. 

Conclusion: We have shown that by using a GWR model to analyze the
relationship between CVM and socioeconomic covariates, we can find the
factors that have the most significant impact on death rates in
different areas of the country. This highlights the need to identify
efficient methods of intervention in order to curb one of the deadliest
groups of diseases on the planet.
[@zelkoGeographicallyWeightedModeling2023]

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::
